Large-scale spatiotemporal variation in vital rates and population dynamics of an alpine bird

### 2024-01-16 Manuscript_Nater_etal.qmd
Text edits and additions to intro, methods and results, and .bib file

---
title: "Large-scale spatiotemporal variation in vital rates and population dynamics of an alpine bird"
format: docx
author: Chloé R. Nater, James A. Martin, Erlend B. Nilsen
editor: visual
bibliography: rypeIDSM_MS2.bib
---

@@

Quantifying the temporal and spatial variation in animal demography is a central theme of ecology and important for directing policy. However, field sampling at large spatial extents for long periods of time is cost prohibitive and often politically untenable. Citizen science programs can alleviate these issues by harnessing stakeholders to increase the spatial and temporal resolution of datasets. Yet, many citizen science programs lack data on individual animals making it difficult to model demography. Integrated population models (IPM) offer promise to use multiple streams of data on populations or multiple aspects of a single dataset to estimate key vital rates for populations or communities. We leveraged a large-scale citizen science monitoring program---the "Hønsefuglportalen"--- to estimate spatial and temporal variation in demography for willow ptarmigan (*Lagopus lagopus*). The core field sampling approach was line-transect distance sampling with age-specific detections---the latter critical to modelling the population dynamics in the IPM. We explored latitudinal and longitudinal variation in recruitment, survival, and density; patterns of density dependence; and the relationship between recruitment and rodent occupancy using an IPM in a Bayesian framework following open science principles. Overall results were consistent with expectations of ecological theory. Slower life history patterns were observed at higher latitudes and altitudes portending differential effects of climate change on ptarmigan across their range. Observed patterns between rodents and recruitment were consistent with the alternative prey hypothesis (APH); however, spatial variation in the directionality and magnitude of effect were substantial. The robustness of the hierarchical structure of our model was instrumental in elucidating these patterns. Our study highlights the power of citizen science, IPMs, and open science principles to estimate spatial and temporal variation in the demography for a culturally important species.

@@

There is growing demand for biodiversity indicators from international unions, national government, local management bodies, and corporate and industry actors. Such indicators ideally represent a wide range of biodiversity's states and functions [e.g. Essential Biodiversity Variables, @pereira2013; @jetz2019], yet the development of suitable indicators for certain attributes, such as species abundance and demography, has been more difficult than for others [@schmeller2018; @waldock2022]. This is at least partially due to challenging requirements regarding spatial scales of useful biodiversity indicators. On one hand, indicators need to be representative at large geographic scales, for example, in the context of countries' reporting towards biodiversity targets [e.g. @feld2009]. On the other hand, indicators also ideally have good spatial resolution as the scales relevant for local-level management and planning are often much smaller [@stevenson2021]. This latter requirement is particularly crucial for (infrastructure) development strategies and for species management and conservation, both of which tend to require knowledge on species abundance and population dynamics (i.e. demographic rates) that is relevant for their county- or municipality-level decision making (+ refs?). Another reason why abundance and population indicators ideally come with good spatial resolution is that there can be substantial amounts of variation in population dynamics and life history of species across space [e.g. @robinson2014; @horswill2019]. This variation that needs to be accounted for to develop successful and sustainable strategies for area use, (harvest) management, and species and biodiversity conservation [@williams2002].

While large-scale, spatially-explicit indicators for species abundance and populations are clearly needed, development and practical implementation are greatly limited due to the reliance of such indicators on the availability of data from large-scale, long-term monitoring programmes [@proencca2017]. Consequently, many countries have been working on setting up, maintaining, and improving such monitoring programmes over the last decades. Many now well-established programmes focus on breeding birds and butterflies, and examples include the North American Breeding Bird Survey (<https://www.pwrc.usgs.gov/bbs/>), UK Butterfly Monitoring Scheme (<https://ukbms.org/>), the Game and Wildlife Conservation Trust Partridge Count Scheme [@aebischer2010], and the Swiss Biodiversity Monitoring (<https://www.biodiversitymonitoring.ch/>). <!-- is there any other one we want to name here specifically? -->

There is a natural trade-off between quality and quantity of data that can be collected in any monitoring programme: collecting high quality data in a structured manner is costly and requires trained specialists, thus limiting the amount of data that can be collected, while participatory monitoring, i.e. the collection of ecological data by members of the public [also called citizen or community science, @fraisl2022], allows to greatly reduce costs and extend spatial and taxonomic scales of monitoring at the expense of data quality and risk of bias [@johnston2023]. Consequently, many large-scale monitoring programmes are often limited to presence-absence or very simple count observations, making them suitable for the development of indicators of species distributions and perhaps population trends, but usually not of abundance, population dynamics, and demographic rates (+ example refs). The exception here are monitoring programmes that succeed in making use of a large number of citizen scientists that have been trained to collect data and record metadata in a structured manner and according to a carefully designed protocol. For example, in the United States hunters participate in the collection of bands and wings from harvested American Woodcock (*Scopolax minor)* to estimate survival and age ratios [@zimmerman2010]. <!-- can we think of 1-2 internationally well-known examples? Check out MAPS (Monitoring of Avian Productivity and Survivorship) and CES (Constant Effort Sites) --> In Norway there is a monitoring programme for terrestrial game bird species called "Hønsefuglportalen" (= "game bird portal", <https://honsefugl.nina.no/Innsyn/en>). It is a line transect survey programme carried out annually in \>120 localities across the country (\>2000 transects) by trained volunteers using pointing dogs. The programme has a well developed protocol for recording bird observations, auxiliary data, and relevant metadata (+ref to protocol?) and established routines for quality control and annual releases of publicly available data via GBIF. As such, it is particularly well suited to become part of a workflow for producing and updating abundance and population indicators on an annual basis.

The line transect data from "Hønsefuglportalen" has been used previously for estimating abundance trends of willow ptarmigan (*Lagopus lagopus*) across Norway [e.g. @bowler2020; @nilsen2020] but large-scale estimation of demographic rates underlying these trends has thus far remained as untapped potential of the dataset. @nilsen2023 recently developed a novel integrated distance sampling model (IDSM) which successfully uses the age of idividuals detected along line transects data coupled with radio-telemetry data to estimate not just abundance, but also survival and recruitment across years. In this study, we adapt the model of @nilsen2023 to run not just on a single site but on all areas with publicly available line transect data from "Hønsefuglportalen" simultaneously. Unlike several previous studies applying integrated models for population dynamics to multiple (sub-) populations separately and comparing results [e.g. @robinson2014; @nater2023], we opt for an approach explicitly integrating across space, thus allowing for sharing of information across locations and -- in effect -- space-for-time substitution [e.g. @horswill2019; @morrison2022]. We then apply the resulting multi-area IDSM to "Hønsefuglportalen" data on willow ptarmigan to estimate population size, age-structure, survival, recruitment, and impacts of small rodent abundance across 41 reporting districts and 15 years (2007-2021) for this popular small-game species. We further embed the modelling workflow in a reproducible, semi-automated pipeline that will greatly facilitate the repeated calculation of abundance and population indicators at different spatial scales as new data are added every year.

@@

The willow ptarmigan is a tetraoid bird with a circumpolar distribution, mainly inhabiting sub-alpine and arctic ecosystems [see e.g. @fuglei2020]. While the species is currently listed as Least Concern (LC) both in the global and Norwegian Red List of Species, it has undergone rather dramatic declines in abundance in Norway since the turn of the 20th century [@hjeljord2022]. While the main reason for the long-term decline in abundance remain unresolved, willow ptarmigan are considered sentinel species that are sensitive to both climate change and land use changes [@Henden2017; @Storch2007]. Moreover, being one of only a handful of bird species that spend the winter in mountain ecosystems in Scandinavia, they are important components of the ecosystem as prey species for the gyrfalcon [@Alastair2020] <!-- do we have other "important" species to add here. Arctic fox? Some other endangered beasts perhaps? -->. The willow ptarmigan has a relatively fast pace of life [@sandercock2005; @Steen1996], and can display substantial spatio-temporal variation in demographic rates [@Bowler2020]. Their population dynamics are characterized by large inter-annual fluctuations in abundance [@hjeljord2022], and previous research has tied these fluctuations to rodent cycle through shared predators [@hagen1952; @Bowler2020]. This tight relationship between the breeding success of ground nesting birds and the rodent cycle is known as the Alternative Prey Hypothesis and has been a central part of Fennoscandian research on grouse population dynamics for many decades [@hagen1952; @Linden1988; @Steen1985]. In addition, spring weather conditions and phenology is known to have considerable effects on breeding success and recruitment rates [@Eriksen2023; @Steen1985]. Across their distributional range, they are iconic species with a high cultural value, partly linked to their popularity as game species. The latter means that information about spatio-temporal variation in demographic rates and population dynamics is particularly important in order to design sustainable harvest strategies. In addition, as a sentinel species the willow ptarmigan is well suited as an indicator species for ecosystem status; in Norway it is included in both the main national biodiversity [Nature Index for Norway, @jakobsson2020, <https://www.naturindeks.no/Indicators/lirype>] and ecosystem [Assessment of the Ecological Condition, @framstad2022] assessments.

@@

The line transect survey data were collected through a structured participatory monitoring program called "Hønsefuglportalen" (<https://honsefugl.nina.no/Innsyn/en>). In August each year, trained volunteer fieldworkers collect observations of willow ptarmigan and other grouse species (rock ptarmigan *Lagopus muta*, black grouse *Lyrurus tetrix*, capercaillie *Tetrao urogallus*) along predefined line transects. To increase the detection probability, each team of fieldworkers use a pointing dog to locate the birds. Each survey team typically consist of at least two persons (one dog handler and one person responsible for following the transect line) and one dog. Often, more than one dog is used for a survey, but only one dog should be used at a time. The transect lines vary in length, but are typically between 1 and 8 km (range: 0.3-16.2 km, median: 3 km). When birds are observed, the exact location of observation is noted, along with its perpendicular distance from the transect line, as well as the age and sex of the birds. An observation typically includes 1 - 12 birds (mean = 5.6), with groups \> 1 typically representing one brood (female and or male with young-of-the-year chicks). When the surveys are conducted in August, the chicks of the year are able to fly but can be distinguished from older birds as they are still of smaller body size. Since 2019, most of the data has been collected using a mobile app tailored made for the monitoring program, where the field workers can register and get access to the transect lines allocated to them by the local organizers. Prior to 2019, data were collected on a dedicated fieldwork form, and entered manually in a web portal afterwards. After field data has been registered, it undergoes several steps of quality control carried out by local stakeholders and personnel from the Norwegian Institute for Nature Research (NINA). Surveys are carried out on both public and private land. After an initial embargo period, all data from public land are published and made freely available as a sampling-event data set on GBIF (refs). The published datasets contain both metadata about the transect surveys (survey date, line transect length and location, study area ID, etc.) as well as bird observation data (species, number of birds of different categories (adult males, adult females, juveniles, and birds of unknown category), perpendicular distance to transect line, exact location, and time of observation). Formally, the data from public land is published as three distinct data sets, one for each of the main public land administrators (Statskog, FeFo and Fjellstyrene, respectively). <!-- I changed "owners" to "administrators", just because to me "public" implies it is not owned by anyone. But if this is wrong, please change it back -->

@@

In this study we used all publicly available data for the period 2007-2021, which included a total of 2225 transects in 41 different reporting districts spanning 9 counties and 50 municipalities. Transects on which no willow ptarmigan were ever observed (i.e. species absence likely due to low habitat suitability) were dropped, resulting in 2077 transects used in the analyses.

@@

The model of @nilsen2023 integrated line transect data with radio-telemetry data from from an ongoing field study of marked willow ptarmigans in Lierne municipality in Central Norway. From 2015 to 2019, around 50 birds were captured in winter (late February or early March) each year and fitted with VHF collars. The marked birds were then monitored on a regular basis until they either i) died, ii) their transmitter's battery stopped working, or iii) we lost contact with the bird for other reasons. For most of the year, the birds were monitored <!-- how were they actually monitored? Can we have a tiny bit more detail here? --> at least once a month. During the breeding and chick-rearing season (May to July) birds were monitored more often, and during December and January we obtained few observations due to challenging field work conditions. A proportion of the birds were harvest annually in the regular recreational harvest, and birds that were harvested were reported as shot to the field personnel. In addition, as the collars had mortality switch, we were also able to locate and retrieve a high proportion of birds that died for natural causes, resulting in a known-fate mark-recapture dataset. The radio-telemetry study is described in detail in @Israelsen2020 and in @Arnekleiv2022.

@@

@nilsen2023 recently developed an integrated distance sampling model (IDSM) which jointly analyses line transect and radio-telemetry data and applied it to willow ptarmigan in the Western part of Lierne municipality in Norway. The model consists of a deterministic population model with two age classes (juveniles and adults) and four data likelihoods: 1) likelihood for observation distances from transect lines for estimating detection probability; 2) likelihood for age-specific counts on transect surveys for estimating numbers of juveniles and adults present; 3) likelihood for juvenile to adult ratios observed at the locality level to provide estimate recruitment rate (as juveniles/adult); and 4) likelihood for known-fate telemetry data to estimate seasonal and annual survival. Below, we describe our new extension of this model to include data from several as opposed to just one. For more detailed information on the single-site model, including tests of model performance, see @nilsen2023.

@@

![Simplified graphical representation of the ptarmigan life cycle with two age classes and the data sources included in the integrated distance sampling model. The pink "x"'s indicate the added dimension for area. Juv\[x,t\] = juveniles in year t. Ad\[x,t\] = adults in year t. R\[x,t\] = recruitment rate in year t. S\[x,t\] = survival probability from year t to t+1. Note that the additional site (=transect) dimension, "j", is omitted for the sake of illustration.](Figures/RypeIDSM_multiArea_structure.png){#fig-model}

@@

The three likelihoods for data resulting from the line transect sampling (observation distances, age-specific counts, and juvenile to adult ratios, see above) were also formulated as spatially explicit, with year- and area-specific distance sampling detection parameters modelled in the same way as survival and recruitment (bar the effect of rodent occupancy, @fig-model). For the known-fate telemetry data (and the seasonal decomposition of survival estimated from it), on the other hand, we did not add an additional area dimension as this data was only available for the Lierne area.

@@

Following model fitting, we calculated posterior distributions for the proportions of variance in survival probabilities, recruitment rates, and detection decay explained by 1) spatial variation ($var_{area}$), 2) temporal variation ($var_{year}$), 3) residual variation ($var_{residual}$), and 4) variation in rodent occupancy ($var_{rodent}$). To obtain the proportion variance explained by each of the component, we divided it by the sum of all other components ($var_{area} + var_{year} + var_{residual} + var_{rodent}$). The spatial, temporal, and residual variance components were defined as the square of the estimated corresponding random effects standard deviation from the model while $var_{rodent}$ was calculated as the variance of all area- and year-specific $\beta_x*rodentOcc_{x,t}$ products.

@@

Reproducibility and ease of repeating analyses was a key focus when developing the multi-area IDSM.To that end, we set up the workflow as a "targets pipeline", implemented through the R package "targets" [@landau2021]. The pipeline contains a variety of options for controlling modelling decisions in the workflow such as the year range of data to consider, the level of spatial aggregation (i.e. reporting district vs. survey locality), whether to model time variation in survival and/or effects of rodent abundance, whether to run MCMC chains sequentially or in parallel, etc. A visual representation of the pipeline is also displayed in @fig-workflow and for more details on pipeline implementation and options, we refer the reader to the GitHub repository: <https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp>.

@@

Only during the most recent four years (2018-2021) has data been collected regularly for all reporting areas included in the analyses. During this period, estimated population densities varied between `r dens.min$post.sum` birds/km^2^ in the area "`r dens.min$area.name`" close to the coast in central Norway to `r dens.max$post.sum` birds/km^2^ in "`r dens.max$area.name`" further south near the Swedish border. In general, population density appeared to be lowest in the Northern parts of Norway and highest in the eastern part of central Norway (@fig-densA). Uncertainty in density estimates was relatively consistent, with a few areas (inlcuding the one with the lowest estimated density, "`r dens.min$area.name`") sticking out by having substantially less precise estimates (@fig-densB). Populations fluctuated substantially over time in any given area (SF "TimeSeries_popDens1.pdf") and some years seemed to be indicative of relatively high (e.g. 2011, 2014, 2018) or low (e.g. 2012, 2015) densities across a large number of areas.

@@

The highest population growth rates were estimated for areas with relatively low population densities across latitudes but we did not find evidence for a strong association between population growth rates and population densities overall ([@fig-paramCorrs] A).

@@

The MCMC chains for many of the area-specific average survival probabilities and recruitment rates, as well as for the global averages (intercepts) for both vital rates, were mixing rather poorly. Despite that, mixing was good and resulting posteriors well defined for the area- and year-specific estimates of survival and recruitment (SF "PostDens_tS_tR.pdf"). There was substantial variation in both vital rates across time (SFs "TimeSeries_pSurv.pdf" and "TimeSeries_rRep.pdf"). In many areas, the years 2011, 2014, and 2018 not only sported high population densities (see above) but were also characterized by both high recruitment and low subsequent survival. The overall low density years 2012 and 2015, conversely, often featured lower recruitment and, in some cases, higher survival. Notably, there were also years with very little spatial synchrony, i.e. very different relative yearly survival probabilities and recruitment rates (e.g. 2010 and 2020 for survival and 2013, 2016, and 2017 for recruitment).

>>>>

---
title: "Large-scale spatiotemporal variation in vital rates and population dynamics of an alpine bird"
format: html
author: Chloé R. Nater, James A. Martin, Erlend B. Nilsen
editor: visual
bibliography: rypeIDSM_MS2.bib
---

@@

Quantifying the temporal and spatial variation in animal demography is a central theme of ecology and important for directing policy. However, field sampling at large spatial extents for long periods of time is cost prohibitive and often politically untenable. Citizen science programs can alleviate these issues by harnessing stakeholders to increase the spatial and temporal resolution of datasets. Yet, most citizen science programs lack data from marked animals making it difficult to model demography. Integrated population models (IPM) offer promise to use multiple streams of data on populations or multiple aspects of a single dataset to estimate key vital rates for populations or communities. We leveraged a large-scale citizen science monitoring program---the "Hønsefuglportalen"--- to estimate spatial and temporal variation in demography for willow ptarmigan (*Lagopus lagopus*). The core field sampling approach was line-transect distance sampling with age-specific detections---the latter critical to modelling the population dynamics in the IPM. We explored latitudinal and longitudinal variation in recruitment, survival, and density; patterns of density dependence; and the relationship between recruitment and rodent occupancy using an IPM in a Bayesian framework following open science principles. Overall results were consistent with expectations of ecological theory. Slower life history patterns were observed at higher latitudes and altitudes portending differential effects of climate change on ptarmigan across their range. Observed patterns between rodents and recruitment were consistent with the alternative prey hypothesis (APH); however, spatial variation in the directionality and magnitude of effect were substantial. The robustness of the hierarchical structure of our model was instrumental in elucidating these patterns. Our study highlights the power of citizen science, IPMs, and open science principles to estimate spatial and temporal variation in the demography for a culturally important species.

@@

There is growing demand for biodiversity indicators from international unions, national government, local management bodies, and corporate and industry actors. Such indicators should ideally represent a wide range of biodiversity's states and functions [e.g. Essential Biodiversity Variables, @pereira2013; @jetz2019], yet the development of suitable indicators for certain attributes, such as species abundance and demography, has been more difficult than for others [@schmeller2018; @waldock2022]. This is at least partially due to challenging requirements regarding spatial scales of useful biodiversity indicators. On one hand, indicators need to be representative at large geographic scales, for example, in the context of countries' reporting towards biodiversity targets [e.g. @feld2009]. On the other hand, indicators also ideally have high spatial resolution, as the scales relevant for local-level management and planning are often much smaller [@stevenson2021]. This latter requirement is particularly crucial for (infrastructure) development strategies and for species management and conservation, both of which tend to require knowledge on species abundance and population dynamics (i.e. demographic rates) that is relevant for their county- or municipality-level decision making (+ refs?). Another reason why abundance and population indicators ideally come with good spatial resolution is that there can be substantial amounts of variation in population dynamics and life history of species across space [e.g. @robinson2014; @horswill2019]. This variation needs to be accounted for to develop successful and sustainable strategies for area use, (harvest) management, and species and biodiversity conservation [@williams2002].

While large-scale, spatially-explicit indicators for species abundance and populations are clearly needed, development and practical implementation are greatly limited due to the reliance of such indicators on the availability of data from large-scale, long-term monitoring programmes [@proencca2017]. Consequently, many countries have been working on setting up, maintaining, and improving such monitoring programmes over the last decades. Many now well-established programmes focus on breeding birds and butterflies, and examples include the North American Breeding Bird Survey (<https://www.pwrc.usgs.gov/bbs/>), the PanEuropean Common Bird Monitoring Scheme (<https://pecbms.info/>[- PECBMS](https://pecbms.info/)), the UK Butterfly Monitoring Scheme ([https://ukbms.org/](#0){.uri style="font-size: 11pt;"}), the Game and Wildlife Conservation Trust Partridge Count Scheme [@aebischer2010], and the Swiss Biodiversity Monitoring ([https://www.biodiversitymonitoring.ch/](#0){.uri style="font-size: 11pt;"}). <!-- is there any other one we want to name here specifically? -->

There is a natural trade-off between quality and quantity of data that can be collected in any monitoring programme: collecting high quality data in a structured manner is costly, requires trained specialists, and depends on at least some top-down control of the program. This often limit the amount of data that can be collected, while participatory monitoring, i.e. the collection of ecological data by members of the public [also called citizen or community science, @fraisl2022], allows to greatly reduce costs and extend spatial and taxonomic scales of monitoring at the expense of data quality and risk of bias [@johnston2023]. Consequently, many large-scale monitoring programmes are often limited to presence-absence or very simple count observations, making them suitable for the development of indicators of species distributions and perhaps population trends, but usually not of abundance, population dynamics, and demographic rates (+ example refs). The exception here are monitoring programmes that succeed in making use of a large number of citizen scientists that have been trained to collect data and record metadata in a structured manner and according to a carefully designed protocol. For example, in the United States hunters participate in the collection of bands and wings from harvested American Woodcock (*Scopolax minor)* to estimate survival and age ratios [@zimmerman2010]. At the European level, the recently established initiative "European Observatory of Wildlife" are offering common field- and analyses protocols and aim to establish a network of "observation points" to monitor wildlife populations at the European level (<https://wildlifeobservatory.org/>). <!-- can we think of 1-2 internationally well-known examples? Check out MAPS (Monitoring of Avian Productivity and Survivorship) and CES (Constant Effort Sites) --> In Norway there is a monitoring programme for terrestrial game bird species called "Hønsefuglportalen" (= "game bird portal", [https://honsefugl.nina.no/Innsyn/en](#0){.uri style="font-size: 11pt;"}). It is a line transect survey programme carried out annually in \>120 localities across the country (\>2000 transects) by trained volunteers using pointing dogs. The programme has a well developed protocol for recording bird observations, auxiliary data, and relevant metadata (+ref to protocol?) and established routines for quality control and annual releases of publicly available data via the Global Biodiversity Information Facility (GBIF). As such, it is particularly well suited to become part of a workflow for producing and updating abundance and population indicators on an annual basis.

The line transect data from "Hønsefuglportalen" has been used previously for estimating abundance trends of willow ptarmigan (*Lagopus lagopus*) across Norway [e.g. @bowler2020; @nilsen2020], and to test relevant ecological hypotheses [@bowler2020; @Breisjøberget_etal_2018]. However, large-scale estimation of demographic rates underlying these trends has thus far remained as untapped potential of the dataset. @nilsen2023 recently developed a novel integrated distance sampling model (IDSM) which successfully uses the age of idividuals detected along line transects data coupled with radio-telemetry data to estimate not just abundance, but also survival and recruitment across years. In this study, we adapt and extend the model of @nilsen2023 to run not just on a single site but on all areas with publicly available line transect data from "Hønsefuglportalen" simultaneously. Unlike several previous studies applying integrated models for population dynamics to multiple (sub-) populations separately and comparing results [e.g. @robinson2014; @nater2023], we opt for an approach explicitly integrating across space, thus allowing for sharing of information across locations and -- in effect -- space-for-time substitution [e.g. @horswill2019; @morrison2022]. We then apply the resulting multi-area IDSM to "Hønsefuglportalen" data on willow ptarmigan to estimate population size, age-structure, survival, recruitment, and impacts of small rodent abundance across 41 reporting districts and 15 years (2007-2021) for this popular small-game species. We further embed the modelling workflow in a reproducible, semi-automated pipeline that will greatly facilitate the repeated calculation of abundance and population indicators at different spatial scales as new data are added every year.

@@

The willow ptarmigan is a tetraoid bird with a circumpolar distribution, mainly inhabiting sub-alpine and arctic ecosystems [see e.g. @fuglei2020]. While the species is currently listed as Least Concern (LC) both in the global and Norwegian Red List of Species, it has undergone rather dramatic declines in abundance in Norway since the turn of the 20th century [@hjeljord2022]. The main reason for the long-term decline in abundance remain unresolved, but the willow ptarmigan are considered sentinel species that are sensitive to both climate change and land use changes [@Henden2017; @Storch2007]. Moreover, being one of only a handful of bird species that spend the winter in mountain ecosystems in Scandinavia, they are important components of the ecosystem as prey species for the gyrfalcon [@Alastair2020] <!-- do we have other "important" species to add here. Arctic fox? Some other endangered beasts perhaps? ERLEND: Well; there are more mammals, as they typically do not migrate long distances to other continents.... I suggest we leave it like this -->. The willow ptarmigan has a relatively fast pace of life [@sandercock2005; @Steen1996], and can display substantial spatio-temporal variation in demographic rates [@Bowler2020]. Their population dynamics are characterized by large inter-annual fluctuations in abundance [@hjeljord2022], and previous research has tied these fluctuations to rodent cycle through shared predators [@hagen1952; @Bowler2020]. This tight relationship between the breeding success of ground nesting birds and the rodent cycle is known as the Alternative Prey Hypothesis (APH) and has been a central part of Fennoscandian research on grouse population dynamics for many decades [@Elton1942; @hagen1952; @Linden1988; @Steen1985]. In addition, spring weather conditions and phenology is known to have considerable effects on breeding success and recruitment rates [@Eriksen2023; @Steen1985]. Across their distributional range, willow ptarmigan are iconic species with a high cultural value, partly linked to their popularity as game species. The latter means that information about spatio-temporal variation in demographic rates and population dynamics is particularly important in order to design sustainable harvest strategies [@Eriksen2023]. In addition, as a sentinel species the willow ptarmigan is well suited as an indicator species for ecosystem status; in Norway it is included in both the main national biodiversity [Nature Index for Norway, @jakobsson2020, <https://www.naturindeks.no/Indicators/lirype>] and ecosystem [Assessment of the Ecological Condition, @framstad2022] assessments.

@@

The line transect survey data were collected through a structured participatory monitoring program called "Hønsefuglportalen" (<https://honsefugl.nina.no/Innsyn/en>). In the first three weeks of August each year, trained volunteer fieldworkers collect observations of willow ptarmigan and other grouse species (rock ptarmigan *Lagopus muta*, black grouse *Lyrurus tetrix*, and capercaillie *Tetrao urogallus*) along predefined line transects. To increase the detection probability, each team of fieldworkers use a pointing dog to locate the birds. Each survey team typically consist of at least two persons (one dog handler and one person responsible for following the transect line) and one dog. Often, more than one dog is used for a survey, but only one dog should be used at a time. The transect lines vary in length, but are typically between 1 and 8 km (range: 0.3-16.2 km, median: 3 km). When birds are observed, the exact location of observation is reported, along with its perpendicular distance from the transect line, as well as the age and sex of the birds. An observation typically includes 1 - 12 birds (mean = 5.6), with groups \> 1 typically representing one brood (female and or male with young-of-the-year chicks). When the surveys are conducted in August, the chicks of the year are able to fly but can be distinguished from older birds as they are still of smaller body size. Since 2019, most of the data has been collected using a mobile app tailored made for the monitoring program, where the field workers can register and get access to the transect lines allocated to them by the local organizers. Prior to 2019, data were collected on a dedicated fieldwork form, and entered manually in a web portal afterwards. After field data has been registered, it undergoes several steps of quality control carried out by local stakeholders and personnel from the Norwegian Institute for Nature Research (NINA). Surveys are carried out on both public and private land. After an initial embargo period, all data from public land are published and made freely available as a sampling-event data set on GBIF (<https://www.gbif.org/sampling-event-data>). The published datasets contain both metadata about the transect surveys (survey date, line transect length and location, study area ID, etc.) as well as bird observation data (species, number of birds of different categories (adult males, adult females, juveniles, and birds of unknown category), perpendicular distance to transect line, exact location, and time of observation). Formally, the data from public land is published as three distinct data sets, one for each of the main public land administrators (Statskog, FeFo and Fjellstyrene, respectively). <!-- I changed "owners" to "administrators", just because to me "public" implies it is not owned by anyone. But if this is wrong, please change it back -->

@@

In this study we used all publicly available data for the period 2007-2021, which included a total of 2225 transects in 41 different reporting districts spanning 9 counties and 50 municipalities. Transects on which no willow ptarmigan were observed during the study period (i.e. species absence likely due to low habitat suitability) were not included. After this initial filtering, a total of 2077 transects were included in the analyses.

@@

The model of @nilsen2023 integrated line transect data with radio-telemetry data from from an ongoing field study of marked willow ptarmigans in Lierne municipality in Central Norway. From 2015 to 2019, around 50 birds were captured in winter (late February or early March) each year and fitted with VHF collars. The marked birds were then monitored on a regular basis until they either i) died, ii) their transmitter's battery stopped working, or iii) we lost contact with the bird for other reasons. For most of the year, the birds were monitored <!-- how were they actually monitored? Can we have a tiny bit more detail here? --> at least once a month by radio triangulation. Most of the fieldwork was conducted from the ground, but to avoid loss of data, the birds were also triangulated from helicopters in May, September and November. During the breeding and chick-rearing season (May to July) birds were monitored more often, and during December and January we obtained few observations due to challenging field work conditions. A proportion of the birds were harvest annually in the regular recreational harvest, and birds that were harvested were reported as shot to the field personnel. In addition, as the collars had mortality switch, we were also able to locate and retrieve a high proportion of birds that died for natural causes, resulting in a known-fate mark-recapture dataset. The radio-telemetry study is described in detail in @Israelsen2020 and in @Arnekleiv2022.

@@

@nilsen2023 recently developed an integrated distance sampling model (IDSM) which jointly analyses line transect and radio-telemetry data and applied it to willow ptarmigan in the Western part of Lierne municipality in Norway. The model consists of a deterministic population model <!-- I think this depends on defenitions. It did contain temporal variation in recruitment, which is in fact environmental stochasticity. It did not contain demographic stochasticity --> with two age classes (juveniles and adults) and four data likelihoods: 1) likelihood for observation distances from transect lines for estimating detection probability; 2) likelihood for age-specific counts on transect surveys for estimating numbers of juveniles and adults present; 3) likelihood for juvenile to adult ratios observed at the locality level to provide estimate recruitment rate (as juveniles/adult); and 4) likelihood for known-fate telemetry data to estimate seasonal and annual survival. Below, we describe our new extension of this model to include data from several as opposed to just one. For more detailed information on the single-site model, including tests of model performance, see @nilsen2023.

@@

![Simplified graphical representation of the ptarmigan life cycle with two age classes and the data sources included in the integrated distance sampling model. The pink "x"'s indicate the added dimension for area. Juv\[x,t\] = juveniles in area x year t. Ad\[x,t\] = adults in area x in year t. R\[x,t\] = recruitment rate in area x in year t. S\[x,t\] = survival probability from year t to t+1 in area x. Note that the additional site (=transect) dimension, "j", is omitted for the sake of illustration.](Figures/RypeIDSM_multiArea_structure.png){#fig-model}

@@

The three likelihoods for data resulting from the line transect sampling (observation distances, age-specific counts, and juvenile to adult ratios, see above) were also formulated as spatially explicit, with year- and area-specific distance sampling detection parameters modelled in the same way as survival and recruitment (except the effect of rodent occupancy, @fig-model). For the known-fate telemetry data (and the seasonal decomposition of survival estimated from it), on the other hand, we did not add an additional area dimension as this data was only available for on study area (the Lierne area).

@@

Following model fitting, we calculated posterior distributions for the proportions of variance in survival probabilities, recruitment rates, and detection decay explained by 1) spatial variation ($var_{area}$), 2) temporal variation ($var_{year}$), 3) residual variation ($var_{residual}$), and 4) variation in rodent occupancy ($var_{rodent}$). To obtain the proportion variance explained by each of the component, we divided it by the sum of all the components ($var_{area} + var_{year} + var_{residual} + var_{rodent}$). The spatial, temporal, and residual variance components were defined as the square of the estimated corresponding random effects standard deviation from the model while $var_{rodent}$ was calculated as the variance of all area- and year-specific $\beta_x*rodentOcc_{x,t}$ products. <!-- We might need to justify the latter a bit, as the rodent variance is calculated different than the other variances -->

@@

Reproducibility and ease of repeating analyses was a key focus when developing the multi-area IDSM. To that end, we set up the workflow as a "targets pipeline", implemented through the R package "targets" [@landau2021]. The pipeline contains a variety of options for controlling modelling decisions in the workflow such as the year range of data to consider, the level of spatial aggregation (i.e. reporting district vs. survey locality), whether to model time variation in survival and/or effects of rodent abundance, whether to run MCMC chains sequentially or in parallel, etc. A visual representation of the pipeline is also displayed in @fig-workflow and for more details on pipeline implementation and options, we refer the reader to the GitHub repository: <https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp>.

@@

Only during the most recent four years (2018-2021) has data been collected regularly for all reporting areas included in the analyses. During this period, estimated population densities varied between `r dens.min$post.sum` birds/km^2^ in the area "`r dens.min$area.name`" close to the coast in central Norway to `r dens.max$post.sum` birds/km^2^ in "`r dens.max$area.name`" further south near the Swedish border. In general, during our study period the population density appeared to be lowest in the Northern parts of Norway and highest in the eastern part of central Norway (@fig-densA). Uncertainty in density estimates was relatively consistent, with a few areas (including the one with the lowest estimated density, "`r dens.min$area.name`") sticking out by having substantially less precise estimates (@fig-densB). Populations fluctuated substantially over time in any given area (SF "TimeSeries_popDens1.pdf") and some years seemed to be indicative of relatively high (e.g. 2011, 2014, 2018) or low (e.g. 2012, 2015) densities across a large number of areas.

@@

The highest population growth rates were estimated for areas with relatively low population densities across latitudes but we did not find evidence for a strong association between population growth rates and population densities overall ([@fig-paramCorrs] A). <!-- Indicate that this is just for the four most recent years. Same for figure below -->

@@

The MCMC chains for many of the area-specific average survival probabilities and recruitment rates, as well as for the global averages for both vital rates, were mixing rather poorly. Despite that, mixing was good and resulting posteriors well defined for the area- and year-specific estimates of survival and recruitment (SF "PostDens_tS_tR.pdf"). There was substantial variation in both vital rates across time (SFs "TimeSeries_pSurv.pdf" and "TimeSeries_rRep.pdf"). In many areas, the years 2011, 2014, and 2018 not only supported high population densities (see above) but were also characterized by both high recruitment and low subsequent survival. <!-- Here and above: I would perhaps consider deleting the reference to 2011 as a "high density year", as there is really a lot of spatial variation in this pattern. I'm not so convinced that is the case in general --> The overall low density years 2012 and 2015, conversely, often featured lower recruitment and, in some cases, higher survival. Notably, there were also years with very little spatial synchrony, i.e. very different relative yearly survival probabilities and recruitment rates (e.g. 2010 and 2020 for survival and 2013, 2016, and 2017 for recruitment).
